---
title: "COVID-19 Vaccination Rates"
author: 'Kiley Hooker (PID: A15441609)'
date: "3/3/2022"
output:
  pdf_document: default
  html_document: default
---

# Read our input data

Here we downloaded the most recently dated "Statewide COVID-19 Vaccines Administered by ZIP Code” CSV file from: https://data.ca.gov/dataset/covid-19-vaccine-progress-dashboard-data-by-zip-code

```{r}
vax <- read.csv("covid19vaccinesbyzipcode_test.csv")
head(vax)
```

```{r}
#ymd(vax$as_of_date)
```

> Q1. What column details the total number of people fully vaccinated? persons_fully_vaccinated

> Q2. What column details the Zip code tabulation area? zip_code_tabulation_area

> Q3. What is the earliest date in this dataset? 2021-01-05

```{r}
vax$as_of_date[ncol(vax)]
```

> Q4. What is the latest date in this dataset? 2022-03-01

```{r}
vax$as_of_date[nrow(vax)]
```


```{r}
skimr::skim(vax)
```

> Q5. How many numeric columns are in this dataset? 9

> Q6. Note that there are “missing values” in the dataset. How many NA values there in the persons_fully_vaccinated column? 18338

```{r}
sum(is.na(vax$persons_fully_vaccinated))
```

> Q7. What percent of persons_fully_vaccinated values are missing (to 2 significant figures)? 17.04%

```{r}
sum(is.na(vax$persons_fully_vaccinated)) / nrow(vax) *100
```

> Q8. [Optional]: Why might this data be missing?




## Working with dates

One of the “character” columns of the data is as_of_date, which contains dates in the Year-Month-Day format.

Dates and times can be annoying to work with at the best of times. However, in R we have the excellent lubridate package, which can make life allot easier. Here is a quick example to get you started:



```{r}
library(lubridate)
age <- today()-ymd("2022-03-03")
age
```

```{r}
time_length(age, "year")
```


First I have to make sure my covid vaccination data date column is in lubridate format.
```{r}
# Specify that we are using the year-month-day format
vax$as_of_date <- ymd(vax$as_of_date)
```

> Q9. How many days have passed since the last update of the dataset? 0

```{r}
today() - vax$as_of_date[1]
```



> Q10. How many unique dates are in the dataset (i.e. how many different dates are detailed)? 61

```{r}
length(unique(vax$as_of_date))
```

# Working with ZIP codes

```{r}
library(zipcodeR)
geocode_zip('92037')
```

```{r}
zip_distance('92037','92109')
```

```{r}
reverse_zipcode(c('92037', "92109") )
```


## Focus on the San Diego Area

Let’s now focus in on the San Diego County area by restricting ourselves first to vax$county == "San Diego" entries. We have two main choices on how to do this. The first using base R the second using the dplyr package:

We have done this the *base R* way quite often like so

```{r}
vax[vax$county == "San Diego", ]
```

An often more convinent way to do this type of "filtering" (a.k.a subsetting) is with the **dplyr**.

```{r}
library(dplyr)
sd <- filter(vax, county == "San Diego")
dim(sd)
```

```{r}
sd.10 <- filter(vax, county == "San Diego" &
                age5_plus_population > 10000)
```

> Q11. How many distinct zip codes are listed for San Diego County? 107

```{r}
length(unique(sd$zip_code_tabulation_area))
```

> Q12. What San Diego County Zip code area has the largest 12 + Population in this dataset? 92154

```{r}
sd[which.max(sd$age12_plus_population), "zip_code_tabulation_area"]
```


Using dplyr select all San Diego “county” entries on “as_of_date” “2022-03-01” and use this for the following questions.

```{r}
sd$as_of_date[nrow(sd)]
```

Let's do this with the most recent date in the data-set (2022-03-01).

> Q13. What is the overall average “Percent of Population Fully Vaccinated” value for all San Diego “County” as of “2022-03-01”? 0.705

```{r}
# Filter to the day
sd.latest <- filter(sd, sd$as_of_date == "2022-03-01")
mean(sd.latest$percent_of_population_fully_vaccinated, na.rm = TRUE)
```

```{r}
summary(sd.latest$percent_of_population_fully_vaccinated, na.rm=TRUE)
```

> Q14. Using either ggplot or base R graphics make a summary figure that shows the distribution of Percent of Population Fully Vaccinated values as of “2022-03-01”?

```{r}
hist(sd.latest$percent_of_population_fully_vaccinated, breaks=30)
```

```{r}
library(ggplot2)
ggplot(sd.latest) + 
  aes(percent_of_population_fully_vaccinated) +
  geom_histogram()
```


## Focus on UCSD/La Jolla

```{r}
ucsd <- filter(sd, zip_code_tabulation_area=="92037")
ucsd[1,]$age5_plus_population
```

> Q15. Using ggplot make a graph of the vaccination rate time course for the 92037 ZIP code area:

```{r}
baseplot <- ggplot(ucsd) +
  aes(as_of_date,
      percent_of_population_fully_vaccinated) +
  geom_point() +
  geom_line(group=1) +
  ylim(c(0,1)) +
  labs(x="Date", y="Percent Vaccinated", title = "Vaccination Rate for CA 92037 (UCSD)")
baseplot
```

## Comparing to similar sized areas

```{r}
# Subset to all CA areas with a population as large as 92037
vax.36 <- filter(vax, age5_plus_population > 36144 &
                as_of_date == "2022-03-01")
head(vax.36)
```

```{r}
ave.36 <- mean(vax.36$percent_of_population_fully_vaccinated, na.rm = T)
ave.36
```

> Q16. Calculate the mean “Percent of Population Fully Vaccinated” for ZIP code areas with a population as large as 92037 (La Jolla) as_of_date “2022-02-22”. Add this as a straight horizontal line to your plot from above with the geom_hline() function?

```{r}
baseplot + 
  geom_hline(yintercept = ave.36, linetype=2, col="red")
```

> Q17. What is the 6 number summary (Min, 1st Qu., Median, Mean, 3rd Qu., and Max) of the “Percent of Population Fully Vaccinated” values for ZIP code areas with a population as large as 92037 (La Jolla) as_of_date “2022-03-01”?

```{r}
summary(vax.36$percent_of_population_fully_vaccinated, na.rm=TRUE)
```


> Q18. Using ggplot generate a histogram of this data.

```{r}
ggplot(vax.36) + 
  aes(percent_of_population_fully_vaccinated) +
  geom_histogram() +
  xlim(c(0,1))
```

> Q19. Is the 92109 and 92040 ZIP code areas above or below the average value you calculated for all these above?

```{r}
vax %>% filter(as_of_date == "2022-03-01") %>%  
  filter(zip_code_tabulation_area=="92109") %>%
  select(percent_of_population_fully_vaccinated)
```

```{r}
vax %>% filter(as_of_date == "2022-03-01") %>%  
  filter(zip_code_tabulation_area=="92040") %>%
  select(percent_of_population_fully_vaccinated)
```

> Q20. Finally make a time course plot of vaccination progress for all areas in the full dataset with a age5_plus_population > 36144.

```{r}
vax.36.all <- filter(vax, age5_plus_population > 36144)

ggplot(vax.36.all) +
  aes(as_of_date,
      percent_of_population_fully_vaccinated, 
      group=zip_code_tabulation_area) +
  geom_line(alpha=0.2, color="blue") +
  ylim(c(0,1)) +
  labs(x="Date", y="Percent Vaccinated",
       title="Vaccination Rate Across California",
       subtitle="Only areas with a population above 36k are shown.") +
  geom_hline(yintercept = ave.36, linetype=2)
```

> Q21. How do you feel about traveling for Spring Break and meeting for in-person class afterwards?

```{r}
sessionInfo()
```



